Internal Gravity Wave Breaking in White Dwarf Binaries

### 2019-04-30 writeups/paper.tex
stash WIP

    \documentclass[
        fleqn,
        usenatbib,
    ]{mnras}
    \usepackage{
        amsmath,
        amssymb,
        newtxtext,
        newtxmath,
        ae, aecompl,
        graphicx
    }
    \usepackage[T1]{fontenc}
    \usepackage[
        labelfont=bf, % caption names are labeled in bold
        font=scriptsize % smaller font for captions
    ]{caption}

    \newcommand*{\rd}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}
    \newcommand*{\pd}[2]{\frac{\partial#1}{\partial#2}}
    \newcommand*{\md}[2]{\frac{\mathrm{D}#1}{\mathrm{D}#2}}
    \newcommand*{\at}[1]{\left.#1\right|}
    \newcommand*{\abs}[1]{\left|#1\right|}
    \newcommand*{\ev}[1]{\langle#1\rangle}
    \newcommand*{\p}[1]{\left(#1\right)}
    \newcommand*{\s}[1]{\left[#1\right]}
    \newcommand*{\z}[1]{\left\{#1\right\}}
    \DeclareMathOperator*{\argmin}{argmin}
    \DeclareMathOperator*{\argmax}{argmax}

@@

@@

We consider a two-dimensional incompressible, isothermal fluid, representative
of degenerate matter in WDs. We neglect temperature variations and assume a
barotropic equation of state as a first approximation. As we are interested in
dynamics far from the center of the WD, we approximate the gravitational field
as uniform. We model the background density stratification as $\rho_0(x, z) =
\rho_0(z) = \rho_0(z=0) e^{-z/H}$ for some reference density $\rho_0(z=0) =
\at{\rho_0(z)}_{z = 0}$.

The Euler equations for an incompressible, barotropic fluid in a uniform
gravitational field are
\begin{subequations}\label{se:nl_orig}
    \begin{align}
        \vec{\nabla} \cdot \vec{u} &= 0,\label{eq:nl_incomp}\\
        \md{\rho}{t} &= 0 ,\label{eq:nl_density}\\
        \md{\vec{u}}{t} + \frac{\vec{\nabla}p}{\rho} + g\hat{z} &=
            0.\label{eq:nl_mom}.
    \end{align}
\end{subequations}
$\md{}{t} = \pd{}{t} + \p{\vec{u} \cdot \vec{\nabla}}$ is the Lagrangian or
material derivative. $\vec{u}, \rho, p$ denote the velocity field, density and
pressure respectively. We denote $-g\hat{z}$ constant gravitational
acceleration. Note that at hydrostatic equilibrium $\pd{}{t} = 0$ we have
$\vec{\nabla}P_0 = -\rho_0 g\hat{z}$ and so $P_0 = \rho_0 gH$. A shear flow
$u_x(x, z, t) = u_x(z)$ is permitted in hydrostatic equilibrium, but we will
initially consider no shear flow in equilibrium.

In practice, it is convenient to introduce coordinate $\Upsilon = \ln
\frac{\rho}{\rho_0}$. This both identically enforces $\rho > 0$ and avoids
numerical issues if $\rho$ is small. We also define reduced pressure $P =
\frac{p}{\rho}$. Then, we may rewrite the second two equations of
\autoref{se:nl_orig} as
\begin{subequations}\label{se:nl_upsilon}
    \begin{align}
        \md{\Upsilon}{t} + u_z \pd{\ln \rho_0}{z} &= 0,\label{eq:nl_up_density}
            \\
        \md{\vec{u}}{t} + \vec{\nabla}P + P\vec{\nabla}\Upsilon + g\hat{z} &= 0.
    \end{align}
\end{subequations}
Note that in the new coordinates, hydrostatic equilibrium corresponds to
$\Upsilon_0 = 1, P_0 = gH$.

@@

In the small perturbation limit, where flow velocities are small compared to the
characteristic space and time scales $\pd{}{t} \gg \vec{u} \cdot \vec{\nabla}$,
we may linearize \autoref{se:nl_upsilon}. We ignore the advective components of the
material derivative and consider small deviations $\Upsilon_1 = \Upsilon -
\Upsilon_0, P_1 = P - P_0$ about hydrostatic equilibrium:
\begin{subequations}\label{se:lin_homo}
    \begin{align}
        \vec{\nabla} \cdot \vec{u} &= 0,\\
        \pd{\Upsilon_1}{t} - \frac{u_{z}}{H} &= 0,\\
        \pd{\vec{u}}{t} + \vec{\nabla}P_1
            + \p{1 + \Upsilon_1} g\hat{z}
            &= 0.
    \end{align}
\end{subequations}

This can be solved to obtain solutions up to undetermined amplitude $A$ the
well-known result \citep{drazin,sutherland0}:
\begin{subequations}\label{se:lin_sol}
    \begin{align}
        u_z\p{x, z, t} &= Ae^{z/2H}e^{i(k_{0x}x + k_{0z}z - \omega_0 t)},\\
        \omega_0^2 &= \frac{N^2k_{0x}^2}{k_{0x}^2 + k_{0z}^2 + \frac{1}{4H^2}},
            \label{eq:disp_rel}
    \end{align}
\end{subequations}
where
\begin{equation}
    N^2 \equiv g^2\p{\rd{\rho}{P} - \at{\pd{\rho}{P}}_{ad}}
        = \frac{g}{H},
\end{equation}
the Brunt-V\"ais\"al\"a frequency is constant

In the weak stratification limit $k_{0z}H \gg 1$, the solution exhibits the
following characteristics:
\begin{itemize}
    \item The amplitude of the wave grows like $e^{z/2H}$. Thus, the
        linear approximation is violated for sufficiently large $z$.

    \item The phase and group velocities can be computed respectively:
        \begin{align}
            \vec{c}_{ph} &=
                \p{\frac{\hat{x}}{k_{0x}} + \frac{\hat{z}}{k_{0z}}}\omega,\\
            \vec{c}_{g} &= \frac{N\p{k_{0x}k_{0z}\hat{x}
                - \p{k_{0z}^2 + \frac{1}{4H^2}}\hat{z}}}
                {\p{k_{0x}^2 + k_{0z}^2 + \frac{1}{4H^2}}^{3/2}} \nonumber\\
                &\approx \frac{\omega k_{0z}}{\abs{\vec{k}_0}^{3/2}}
                    \p{k_{0x}\hat{x} - k_{0z}\hat{z}}.
        \end{align}
        We recover the usual result $\vec{c}_{ph} \cdot \vec{c}_g = 0$
        (e.g.\  \citep{drazin,sutherland1}). Note that an IGW transporting energy
        and momentum upwards $c_{g, z} > 0$ has $k_{0z} < 0$.

    \item The time-averaged total $x$-momentum flux in the $\hat{z}$ direction
        can be computed. Since the linear solution is separable as $f(x, z, t) =
        f(z)e^{i(k_{0x}x - \omega_0 t)}$, $x$ averaging and time averaging are
        equivalent, so we may write
        \begin{align}
            S &\equiv \ev{\rho u_x u_z}_x \equiv
                \frac{1}{L_x}\int_0^{L_x}\limits \rho u_x u_z\;\mathrm{d}x,
                    \label{eq:S_def},\\
                &\approx -\frac{A^2}{2}\rho_0(z=0)\frac{k_{0z}}{k_{0x}},
                    \label{eq:S_lin}
        \end{align}
        where $\ev{\dots}_x$ denotes $x$-averaging.
\end{itemize}

\subsection{Wave Generation}

To model a continuous IGW wave train excited deep in the WD interior propagating
towards the surface, we use a volumetric forcing term to excite IGW near the
bottom of the simulation domain\footnote{Interfacial forcing at the bottom
boundary incurs strong stability limits, as we use a Chebyshev polynomial basis
along the $z$ axis in our spectral method which has very small grid spacing at
the boundaries.}. Our forcing excites both IGWs propagating upwards, imitating a
wave tidally excited deeper in the WD, and downwards, which are damped away by
the damping layers described in \autoref{ss:damping}.

As not to interfere with the incompressibility constraint, we force the system
on the density equation. This constitutes replacing \autoref{eq:nl_up_density}
with
\begin{equation}
    \md{\Upsilon}{t} + u_z\pd{\ln \rho_0}{z}
        = Fe^{-\frac{(z - z_0)^2}{2\sigma^2}}
            \cos \p{k_{0x}x - \omega_0 t}.\label{eq:vol_drive}
\end{equation}
Using a narrow Gaussian profile excites a broad $k_z$ wavenumber spectrum, and
only the $k_{0z}$ satisfying dispersion relation \autoref{eq:disp_rel} for the
given $k_{0x}, \omega_0(k_{0x}, k_{0z})$ will propagate.

In the linearized system, the effect of this forcing can be solved analytically
up to good accuracy: we first approximate the driving term using
$e^{\frac{-(z - z_0)^2}{2\sigma^2}} \approx \sqrt{2\pi \sigma^2}\delta(z -
z_0)$, the $\sigma \to 0$ limit\footnote{In practice, $\sigma$ must be large
enough to be numerically resolved by the spectral code.}. This system is solved
exactly by matching the two homogeneous solutions above and below $z_0$. We may
then approximately relax the solution to nonzero $\sigma$: an extra factor of
$e^{-\frac{(k_{0z}\sigma)^2}{2}}$ arises compared to the $\delta$-function
solution (evaluating the Fourier Transform of $e^{-\frac{(z -
z_0)^2}{2\sigma^2}}$ at $k_z = k_{0z}$), and we obtain
\begin{align}
    u_{1z}&(x, z, t) ={} \frac{Fgk_{0x}^2}{\omega_0^2}
        \frac{1}{2ik_{0z}}\frac{e^{\frac{-(k_{0z}\sigma)^2}{2}}}
        {\sqrt{2\pi\sigma^2}} \times\nonumber\\
        &{}\begin{cases}
        e^{\frac{z - z_0}{2H}}e^{i\p{k_{0x}x + k_{0z}(z - z_0) - \omega_0 t
            + \frac{1}{2k_{0z}H}}}
            & z > z_0\\
        e^{\frac{z - z_0}{2H}}e^{i\p{k_{0x}x - k_{0z}(z - z_0) - \omega_0 t
            + \frac{1}{2k_{0z}H}}}
            & z < z_0\\
    \end{cases}.\label{eq:uz_lin}
\end{align}
The $z > z_0$ solution models an upwards-propagating IGW wavetrain inbound on
the simulation domain from below. The solution for $u_{1x}(x, z, t)$ can be
obtained analytically from \autoref{eq:nl_incomp}.

\subsection{Wave Breaking}

In developing the results above, we have neglected advective terms $\pd{}{t} \gg
\vec{u} \cdot \vec{\nabla}$. However, since $\abs{\vec{u}} \propto e^{z/2H}$,
any nonzero amplitude IGW will eventually violate the linearity criterion. This
regime is relevant in practice since the IGW may grow to nonlinear amplitudes
before it reaches the surface of the WD \citep{fullerI,fullerII}. We can coarsely
estimate the height at which this happens by when the Lagrangian displacement
$\vec{\xi}$ of a fluid parcel satisfies
\begin{equation}
    \vec{\xi} \cdot \vec{k} \gtrsim 1.\label{eq:nl}
\end{equation}

The general understanding of nonlinear IGW breaking is that the waves'
interactions with the mean flow of the fluid transfer horizontal momentum from
the waves into the fluid; such a process has been conjectured to be responsible
for tidal synchronization in stellar binaries \citep{zahn75,gn89} as well as the
quasi-biennial oscillation \citep{lindzen_qbo}. The details of this process have
been laid out in a few key papers: IGWs are globally unstable to resonant
three-wave interactions \citep{drazin}, causing energy transfer out of the IGW
to daughter modes. These steeper daughter modes facilitate wave breaking,
depositing horizontal momentum in the fluid's mean flow \citep{klostermeyer}.
Once the mean flow reaches the horizontal phase velocity of the parent IGW, a
critical layer forms at which the frequency of the IGW is Doppler-shifted to
zero. The interaction of the parent IGW with the mean flow was first studied in
the inviscid, linear regime in \citep{booker_bretherton}, which found nearly
complete absorption. This result was reproduced with nonzero viscosity
\citep{hazel}, but weakly nonlinear theory \citep{brown_stewartson} and fully
nonlinear simulations \citep{winters1994} suggest that nonlinear effects can
induce reflection.

An estimate of where these mean flow effects become important can be made using
the wave-induced mean flow \citep{eliassen_palm_cite,sutherland0}:
\begin{equation}
     \ev{u_x} = \bar{U}_0 \equiv \frac{1}{L_x}
        \int\limits u_x\;\mathrm{d}x = \frac{\ev{u_xu_z}_x}{c_{g,z}},
        \label{eq:mean_flow}
\end{equation}
where $\ev{\dots}_x$ denotes horizontal averaging. Setting $\bar{U}_0 = c_{ph,
x}$ the critical layer condition, we recover $\frac{u_xk_x}{\omega} = \xi_x k_x
\sim 1$, roughly equivalent to the nonlinearity criterion \autoref{eq:nl}.

@@

The time integration uses a split implicit-explicit third-order scheme where
certain terms are treated implicitly and the remaining terms are treated
explicitly. A third-order, four-stage DIRK-ERK scheme \citep{ascher} is used
with adaptive timesteps computed from advective Courant-Friedrichs-Lewy (CFL)
time. Specifically, we use $\Delta t = 0.7 \min(\Delta x / u_x,\Delta z / u_z)$,
where the minimum is taken over every grid point in the domain and $\Delta
x,\Delta z$ are the grid spacings in the $x$ and $z$ directions respectively.

The physics of our simulation is fixed by six parameters: $L_x, L_z$, and
$k_{0x}, \omega_0, F, \nu$. We describe our choices for these parameters below:
\begin{itemize}
    \item $L_z$: Since the wave amplitude grows $\sim e^{z/2H}$, we choose $L_z
        = 10H$ to give $\sim e^4$ apmlitude growth between the damping zones.

    \item $L_x$: We want similar grid spacing $\frac{L_x}{N_x} \sim
        \frac{L_z}{N_z}$. For computational savings, we fix $N_z = 4N_x$.
        Expecting wave breaking to be mostly in the $z$ direction, we use $L_x =
        4H$.

    \item $k_{0x}$: Astrophysical IGWs in stars are generally excited by the $l
        = 2$ component of the tidal potential, for which $k_{\perp} \sim
        \frac{1}{R}$ where $R$ is the radius of the star. To best emulate this,
        we use $k_{0x} = \frac{2\pi}{L_x}$ the smallest permitted wavenumber
        permitted by periodic boundary conditions.

    \item $\omega_0$: We choose $\omega_0$ by evaluating dispersion relation
        $\omega_0\p{k_{0x}, k_{0z}}$ for some desired $k_{0z}$.

        Astrophysical IGWs also generally satisfy $\omega \ll N$, or
        equivalently $\frac{k_r}{k_{\perp}} \sim k_rR \gg 1$. Since $H \lesssim
        R$, we aim to study $k_{0z}H \gg 1$. However, to satisfy this well is
        computationally expensive: $L_z \gg H$ is required to give waves ample
        room to grow within the simulation domain, but we want $k_{0z}$ to be
        sufficiently separated from the grid spacing $\sim L_z / N_z$ that
        nonlinear effects can be seen in later simulations (see
        \autoref{ss:nl_ns}). Thus, $N_z$ sets how well we can satisfy $k_{0z}H
        \gg 1$; we choose $k_{0z}H = 1$.

    \item $F$: We first choose $F$ forcing strength such that $\vec{\xi} \cdot
        \vec{k} \ll 1$ is satisfied everywhere throughout the domain. This
        becomes a constraint on $F$ by using \autoref{eq:uz_lin} and
        $\abs{\vec{u}} \propto e^{z/2H}$.

>>>>

    \documentclass[
        fleqn,
        usenatbib,
    ]{mnras}
    \usepackage{
        amsmath,
        amssymb,
        newtxtext,
        newtxmath,
        ae, aecompl,
        graphicx,
        booktabs,
    }
    \usepackage[T1]{fontenc}
    \usepackage[
        labelfont=bf, % caption names are labeled in bold
        font=scriptsize % smaller font for captions
    ]{caption}

    \newcommand*{\rd}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}
    \newcommand*{\pd}[2]{\frac{\partial#1}{\partial#2}}
    \newcommand*{\md}[2]{\frac{\mathrm{D}#1}{\mathrm{D}#2}}
    \newcommand*{\at}[1]{\left.#1\right|}
    \newcommand*{\abs}[1]{\left|#1\right|}
    \newcommand*{\ev}[1]{\langle#1\rangle}
    \newcommand*{\p}[1]{\left(#1\right)}
    \newcommand*{\s}[1]{\left[#1\right]}
    \newcommand*{\z}[1]{\left\{#1\right\}}
    \DeclareMathOperator*{\argmin}{argmin}
    \DeclareMathOperator*{\argmax}{argmax}
    \DeclareMathOperator*{\med}{med}

@@

[Below copied from proposal, will rewrite]

@@

We consider a two-dimensional incompressible, isothermal fluid, representative
of degenerate matter in WDs. We furthermore neglect temperature variations and
assume a barotropic equation of state $P(\rho, T) = P(\rho)$ as a first
approximation. As we are interested in dynamics far from the center of the WD,
we approximate the gravitational field as uniform. We model the background
density stratification as $\rho_0(x, z) = \rho_0(z) = \rho_0(z=0) e^{-z/H}$ for
some reference density $\rho_0(z=0) = \at{\rho_0(z)}_{z = 0}$.

The Euler equations for an incompressible, barotropic fluid in a uniform
gravitational field are
\begin{subequations}\label{se:nl_orig}
    \begin{align}
        \vec{\nabla} \cdot \vec{u}_1 &= 0,\label{eq:nl_incomp}\\
        \md{\rho}{t} &= 0 ,\label{eq:nl_density}\\
        \md{\vec{u}_1}{t} + \frac{\vec{\nabla}p}{\rho} + g\hat{z} &=
            0.\label{eq:nl_mom}.
    \end{align}
\end{subequations}
$\md{}{t} = \pd{}{t} + \p{\vec{u}_1 \cdot \vec{\nabla}}$ is the Lagrangian or
material derivative, and $\vec{u}_1, \rho, p$ denote the velocity field, density
and pressure respectively. We denote $-g\hat{z}$ constant gravitational
acceleration. Note that at hydrostatic equilibrium $\pd{}{t} = 0$ we have
$\vec{\nabla}P_0 = -\rho_0 g\hat{z}$ and so $P_0 = \rho_0 gH$. A shear flow of
form $\vec{u}_1 = u_{1x}(x, z, t)\hat{x} = u_{1x}(z)\hat{x}$ is permitted in hydrostatic
equilibrium, but we will assume no background shear flow. Physically, this
assumption corresponds to a non-rotating star, or going to the corotating frame
of a rigidly rotating star.

In practice, it is convenient to introduce coordinate $\Upsilon = \ln
\frac{\rho}{\rho_0}$. This both identically enforces $\rho > 0$ and avoids
numerical issues in the $\frac{\vec{\nabla}p}{\rho}$ term if $\rho$ is small. We
also define reduced pressure $P = \frac{p}{\rho}$. Then, we may rewrite the
second two equations of \autoref{se:nl_orig} as
\begin{subequations}\label{se:nl_upsilon}
    \begin{align}
        \md{\Upsilon}{t} + u_{1z} \pd{\ln \rho_0}{z} &= 0
            ,\label{eq:nl_up_density} \\
        \md{\vec{u}_1}{t} + \vec{\nabla}P + P\vec{\nabla}\Upsilon + g\hat{z} &= 0.
    \end{align}
\end{subequations}
Note that in the new coordinates, hydrostatic equilibrium corresponds to
$\Upsilon_0 = 1, P_0 = gH$.

@@

In the small perturbation limit, where flow velocities are small compared to the
characteristic space and time scales $\pd{}{t} \gg \vec{u}_1 \cdot \vec{\nabla}$,
we may linearize \autoref{se:nl_upsilon}. We ignore the advective components of the
material derivative and consider small deviations $\Upsilon_1 = \Upsilon -
\Upsilon_0, P_1 = P - P_0$ about hydrostatic equilibrium:
\begin{subequations}\label{se:lin_homo}
    \begin{align}
        \vec{\nabla} \cdot \vec{u}_1 &= 0,\\
        \pd{\Upsilon_1}{t} - \frac{u_{z}}{H} &= 0,\\
        \pd{\vec{u}_1}{t} + \vec{\nabla}P_1
            + \p{1 + \Upsilon_1} g\hat{z}
            &= 0.
    \end{align}
\end{subequations}

This can be solved to obtain solutions up to undetermined amplitude $A$ the
well-known result \citep{drazin,sutherland0}:
\begin{equation}
    u_{1z}\p{x, z, t} = Ae^{z/2H}e^{i\p{k_{1x}x - k_{1z}z - \omega_1 t}},
        \label{eq:k0z_sign}
\end{equation}
where $\omega_1$ satisfies dispersion relation
\begin{equation}
    \omega_1^2 = \frac{N^2k_{1x}^2}{k_{1x}^2 + k_{1z}^2 + \frac{1}{4H^2}},
        \label{eq:disp_rel}
\end{equation}
where
\begin{equation}
    N^2 \equiv g^2\p{\rd{\rho}{P} - \at{\pd{\rho}{P}}_{ad}} = \frac{g}{H},
\end{equation}
the Brunt-V\"ais\"al\"a frequency is constant. Our sign convention is such that
a wave with $k_{1x}, k_{1z} > 0$ has positive vertical group velocity (see
\autoref{eq:vg}).

In the weak stratification limit $k_{1z}H \gg 1$, the solution exhibits the
following characteristics:
\begin{itemize}
    \item The amplitude of the wave grows with $z$ as $e^{z/2H}$. Thus, the
        linear approximation is always violated for sufficiently large $z$.

    \item The phase and group velocities can be computed respectively:
        \begin{align}
            \vec{c}_{ph} &=
                \p{k_{1x}\hat{x} - k_{1z}\hat{z}}\frac{\omega_1}
                {k_{1x}^2 + k_{1z}^2 + \frac{1}{4H^2}},\\
            \vec{c}_{g} &= N\frac{\p{k_{1z}^2 + \frac{1}{4H^2}}\hat{x}
                + \p{k_{1x}k_{1z}\hat{z}}}
                {\p{k_{1x}^2 + k_{1z}^2 + \frac{1}{4H^2}}^{3/2}}.\label{eq:vg}
        \end{align}
        We note $\vec{c}_{ph} \cdot \vec{c}_g = +
        \mathcal{O}\p{\frac{1}{\p{k_{1z}H}^{2}}} \approx 0$, the usual result
        (e.g.\  \citep{drazin,sutherland1}). Note again that $c_{gz} > 0$ for
        $k_{1x}, k_{1z} > 0$, a consequence of our sign choice in
        \autoref{eq:k0z_sign}.

    \item The time-averaged total $x$-momentum flux carried in the $\hat{z}$
        direction can be computed. Since the linear solution is separable as
        $f(x, z, t) = f(z)e^{i(k_{1x}x - \omega_1 t)}$, $x$ averaging and time
        averaging are equivalent, so we may write
        \begin{align}
            S &\equiv \ev{\rho u_{1x} u_{1z}}_x \equiv
                \frac{1}{L_x}\int_0^{L_x}\limits \rho u_{1x} u_{1z}\;\mathrm{d}x,
                    \label{eq:S_def},\\
                &\approx -\frac{A^2}{2}\rho_0(z=0)\frac{k_{1z}}{k_{1x}},
                    \label{eq:S_lin}
        \end{align}
        where $\ev{\dots}_x$ denotes $x$-averaging.
\end{itemize}

\subsection{Wave Generation: Analytical Analysis}

Since we are only concerned with the behavior of the wave at high $z$ where wave
breaking occurs, we generate waves in the most numerically convenient way. As
long as physically represesentative IGW are generated and wave dynamics are only
examined far from the forcing region, the exact excitation mechanism is not
important for our conclusions.

To model continuous excitation of IGWs deep in the WD interior propagating
towards the surface, we use a volumetric forcing term to excite IGW near the
bottom of the simulation domain\footnote{Interfacial forcing is numerically
unstable when using a Chebyshev polynomial basis in the $z$ direction and
impossible when using a Fourier basis.}. Our forcing excites both IGWs
propagating upwards, imitating a wave tidally excited deeper in the WD, and
downwards, which are damped away by the damping layers described in
\autoref{ss:damping}.

As not to interfere with the incompressibility constraint, we force the system
on the density equation. This constitutes replacing \autoref{eq:nl_up_density}
with
\begin{equation}
    \md{\Upsilon}{t} + u_{1z}\pd{\ln \rho_0}{z}
        = Fe^{-\frac{(z - z_0)^2}{2\sigma^2}}
            \cos \p{k_{1x}x - \omega_1 t}.\label{eq:vol_drive}
\end{equation}
Using a narrow Gaussian profile excites a broad $k_z$ wavenumber spectrum, and
only the $k_{1z}$ satisfying dispersion relation \autoref{eq:disp_rel} for the
given $k_{1x}, \omega_1(k_{1x}, k_{1z})$ will propagate.

In the linearized system, the effect of this forcing can be solved analytically
up to good accuracy: we first approximate the driving term using
$e^{\frac{-(z - z_0)^2}{2\sigma^2}} \approx \sqrt{2\pi \sigma^2}\delta(z -
z_0)$, the $\sigma \to 0$ limit\footnote{In practice, $\sigma$ must be large
enough to be numerically resolved by the spectral code.}. This system is solved
exactly by matching the two homogeneous solutions above and below $z_0$. We may
then approximately relax the solution to nonzero $\sigma$: an extra factor of
$e^{-\frac{(k_{1z}\sigma)^2}{2}}$ arises compared to the $\delta$-function
solution (evaluating the Fourier Transform of $e^{-\frac{(z -
z_0)^2}{2\sigma^2}}$ at $k_z = k_{1z}$), and we obtain
\begin{align}
    u_{z}&(x, z, t) ={} \frac{Fgk_{1x}^2}{\omega_1^2}
        \frac{1}{2ik_{1z}}\frac{e^{\frac{-(k_{1z}\sigma)^2}{2}}}
        {\sqrt{2\pi\sigma^2}} \times\nonumber\\
        &{}\begin{cases}
        e^{\frac{z - z_0}{2H}}e^{i\p{k_{1x}x + k_{1z}(z - z_0) - \omega_1 t
            + \frac{1}{2k_{1z}H}}}
            & z > z_0\\
        e^{\frac{z - z_0}{2H}}e^{i\p{k_{1x}x - k_{1z}(z - z_0) - \omega_1 t
            + \frac{1}{2k_{1z}H}}}
            & z < z_0\\
    \end{cases}.\label{eq:uz_lin}
\end{align}
The $z > z_0$ solution models an upwards-propagating IGW wavetrain inbound on
the simulation domain from below. The solution for $u_{1x}(x, z, t)$ can be
obtained analytically from incompressibility \autoref{eq:nl_incomp}.

\subsection{Wave Breaking Location}\label{ss:wave_breaking}

In developing the results above, we have neglected advective terms $\pd{}{t} \gg
\vec{u}_1 \cdot \vec{\nabla}$. However, since $\abs{\vec{u}_1} \propto
e^{z/2H}$, any nonzero amplitude IGW will eventually violate the linearity
criterion. This regime is relevant in practice since the IGW may grow to
nonlinear amplitudes before it reaches the surface of the WD
\citep{fullerI,fullerII}. We can coarsely estimate the height at which this
happens by when the Lagrangian displacement $\vec{\xi} =
\frac{\vec{u}_1}{-i\omega}$ of a fluid parcel satisfies
\begin{equation}
    \vec{\xi} \cdot \vec{k}_1 \gtrsim 1.\label{eq:nl}
\end{equation}

Another useful result quantifies the wave-induced mean flow
\citep{eliassen_palm_cite,sutherland0}:
\begin{equation}
     \ev{u_{1x}} \equiv \bar{U} = \frac{1}{L_x}
        \int\limits u_{1x}\;\mathrm{d}x = \frac{\ev{u_{1x}u_{1z}}_x}{c_{g,z}},
        \label{eq:mean_flow}
\end{equation}
This formula is analogous to Stokes' drift for surface waves and can be derived
by considering the propagation of $S$ (\autoref{eq:S_def}) into a medium at
rest. Solving for $\bar{U} = c_{ph, x}$ the critical layer condition (see
\autoref{ss:crit_layer} immediately below), we recover
$\frac{u_{1x}k_{1x}}{\omega} = \xi_x k_{1x} \sim 1$, roughly equivalent to the
nonlinearity criterion \autoref{eq:nl}. This condition predicts where the wave's
self-acceleration alone is sufficient to induce critical layer absorption.

\subsection{Formation and Dynamics of Critical Layer}\label{ss:crit_layer}

[TODO rewrite more clearly, it's just copy pasted together right now]

The general understanding of nonlinear IGW breaking is that the waves'
interactions with the mean flow of the fluid transfer horizontal momentum from
the waves into the fluid; such a process has been conjectured to be responsible
for tidal synchronization in stellar binaries \citep{zahn75,gn89} as well as the
quasi-biennial oscillation \citep{lindzen_qbo}. The details of this process have
been laid out in a few key papers: IGWs are globally unstable to resonant
three-wave interactions \citep{drazin}, causing energy transfer out of the IGW
to daughter modes. These steeper daughter modes facilitate wave breaking,
depositing horizontal momentum in the fluid's mean flow \citep{klostermeyer}.

In accordance with literature and the simulation setup, we will assume momentum
redistribution is inefficient (see \autoref{ss:disp} for analysis of this
assumption). The mean flow will then grow in place until it reaches the
horizontal phase velocity of the parent IGW\@. Then, a critical layer forms at
which the frequency of the incident IGW is Doppler-shifted to zero. The
interaction of the parent IGW with the mean flow was first studied in the
inviscid, linear regime in \citep{booker_bretherton}, which found nearly
complete absorption. This result was reproduced with nonzero viscosity
\citep{hazel}, but weakly nonlinear theory \citep{brown_stewartson} and fully
nonlinear simulations \citep{winters1994} suggest that nonlinear effects can
induce reflection.

A purely horizontal shear flow $\bar{U}(z) \hat{x}$ can be seen in
\autoref{se:nl_orig} to have the effect of modifying time derivatives
$\partial_t$ to their frequency in the comoving frame of the fluid $\partial_t -
\bar{U}(z)\partial_x$. For a critical value $\omega_1 - \bar{U}_c k_{1x} = 0$,
the frequency of the linear wave in the fluid's frame of reference vanishes and
critical behavior is observed. In a linear \citep{booker_bretherton} or viscous
\citep{hazel} theory, the incident wave has amplitude reflection and
transmission coefficients
\begin{align}
    \mathrm{Ri} &\equiv \at{\frac{N^2}{\p{\pd{\bar{U}}{z}}^2}}_{z_c},
        \label{eq:ri_def}\\
    \mathcal{R} &= e^{-2\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}}, &
    \mathcal{T} &= e^{-\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}},
        \label{eq:crit_coeffs}
\end{align}
where we have defined local Richardson number Ri at the critical layer $z_c:
\bar{U}(z_c) = \frac{\omega_1}{k_{1x}}$. In the $\mathrm{Ri} \gg 1$ limit,
$\mathcal{R}, \mathcal{T} \ll 1$ and the incident wave is completely absorbed to
good approximation.

When the fluid absorbs the incident wave, it absorbs the incident horizontal
momentum flux as well, which is converted into additional horizontal momentum of
the shear flow. Since the shear flow cannot exceed $\bar{U}_c$ the horizontal
phase velocity of the incident wave, the critical layer must instead propagate
downwards to accommodate the incident momentum. The total horizontal momentum of
the shear flow then obeys conservation equation
\begin{equation}
    \pd{}{t}\int\limits_0^{L_z} \rho(z) \bar{U}(z, t)\;\mathrm{d}z
        \Delta S = 0.
\end{equation}
We define $\Delta S$ to be the change in flux across the critical layer, or
equivalently the negative of the absorbed flux. Treating $\bar{U}(z > z_c) =
\bar{U}_c, \bar{U}(z < z_c) = 0$ gives us exactly
\begin{equation}
    \rho(z_c) \bar{U}_c\pd{z_c}{t} = \Delta S(z_c(t)).\label{eq:zc_anal}
\end{equation}
$\Delta S(z_c)$ denotes the change in $S$ across $z_c$. For constant $\Delta S$
in time and $\rho \approx \rho_0$, this has analytical solution
\begin{equation}
    z_c(t) = -H\ln t - H\ln \frac{H\rho_0(z = 0)c_{ph, x}}{-\Delta S}
        .\label{eq:zc_sol}
\end{equation}
Note that constant $-\Delta S$ corresponds physically to the picture where
constant $S$ carried by the IGWs, given by \autoref{eq:S_lin}, is fully absorbed
at the critical layer, in which case $\Delta S = -S$.

@@

The geometry of our simulation domain is fixed by three parameters: $L_x, L_z$,
and $z_0$ the forcing location. We use $z_0 = 0.2L_z$ to force sufficiently far
from the lower damping zone and permit sufficient room for the upwards-moving
wave to grow at $\propto e^{z/2H}$. We choose $L_z = 10H$ to give $\sim e^3$
apmlitude growth between the damping zones. Finally, we want similar grid
spacing $\frac{L_x}{N_x} \sim \frac{L_z}{N_z}$. For computational savings, we
fix $N_z = 4N_x$. We use $L_x = 4H = \frac{2}{5}L_z$.

The time integration uses a split implicit-explicit third-order scheme where
certain terms are treated implicitly and the remaining terms are treated
explicitly. A third-order, four-stage DIRK-ERK scheme \citep{ascher} is used
with adaptive timesteps computed from advective Courant-Friedrichs-Lewy (CFL)
time. Specifically, we use $\Delta t = 0.7 \min(\Delta x / u_x,\Delta z / u_{1z})$,
where the minimum is taken over every grid point in the domain and $\Delta
x,\Delta z$ are the grid spacings in the $x$ and $z$ directions respectively.

The physics of our simulation is fixed by four parameters: $k_{1x}, \omega_1, F,
\nu$. We describe our choices for these parameters below:
\begin{itemize}
    \item $k_{1x}$: Astrophysical IGWs in stars are generally excited by the $l
        = 2$ component of the tidal potential, for which $k_{\perp} \sim
        \frac{1}{R}$ where $R$ is the radius of the star. To best emulate this,
        we use $k_{1x} = \frac{2\pi}{L_x}$ the smallest permitted wavenumber
        permitted by periodic boundary conditions.

    \item $\omega_1$: We choose $\omega_1$ by evaluating dispersion relation
        $\omega_1\p{k_{1x}, k_{1z}}$ for some desired $k_{1z}$.

        Astrophysical IGWs also generally satisfy $\omega \ll N$, or
        equivalently $\frac{k_r}{k_{\perp}} \sim k_rR \gg 1$. Since $H \lesssim
        R$, we aim to study $k_{1z}H \gg 1$. However, to satisfy this well is
        computationally expensive: $L_z \gg H$ is required to give waves ample
        room to grow within the simulation domain, but we want $k_{1z}$ to be
        sufficiently separated from the grid spacing $\sim L_z / N_z$ that
        nonlinear effects can be seen in later simulations (see
        \autoref{ss:nl_ns}). Thus, $N_z$ sets how well we can satisfy $k_{1z}H
        \gg 1$; we choose $k_{1z}H = 1$.

    \item $F$: We first choose $F$ forcing strength such that $\vec{\xi} \cdot
        \vec{k}_1 \ll 1$ is satisfied everywhere throughout the domain. This
        constraints $F$ by \autoref{eq:uz_lin}.

    \item $\nu$: Nonlinear effects transfer wave energy from $\vec{k}_{1}$ to
        smaller scales. Since well-resolved simulations using spectral methods
        have no inherent numerical viscosity, energy will accumulate at grid
        scales in the absence of artificial dissipation. We introduce
        dissipation parameter $\nu$ used for both artificial viscosity and artificial
        diffusivity. We ensure our equations of motion continue to conserve
        horizontal momentum (see \autoref{ss:strat_impl} for details). Under
        weak forcing, we may set $\nu = 0$, since energy transfer out of the
        parent $\vec{k}_1$ mode is negligible.
\end{itemize}
